\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2012-12-17 flux.tex
Acknowledge GISAID laboratories.

Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations proceding in a random direction.
However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
This result indicates that forward movement along antigenic dimension 1 is an advantageous trait and is promoted by natural selection \cite{Bedford12}.
The rate of diffusion of persistent lineages is substantially greater in A/H3N2 than in A/H1N1, B/Vic or B/Yam.
For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drifts along dimension 1.

%%% accumulation %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/accumulation}
	\caption{\textbf{Rate of antigenic diffusion of influenza lineages categorized by time back from sampling.}
	(A) Rate of diffusion in terms of antigenic units per year only along antigenic dimension 1.
	(B) Rate of diffusion in terms of antigenic units per year along both antigenic dimensions 1 and 2.
	In the diffusion model, the antigenic location of each virus sample can be calculcated back through time.
	Here, when the $x$-axis is 1, we show the rate of diffusion 1 year back from from the date of sampling averaged across all virus samples.
	When the $x$-axis is close to 0, the $y$-axis shows the rate of diffusion near tips of the phylogeny, and when the $x$-axis is large, the $y$-axis shows the rate of diffusion along the trunk of the phylogeny.
	} 
	\label{accumulation} 
\end{figure}

@@

We suggest that this finding could be due to the pleiotropic effects associated with antigenic change, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantagous, while other antigenic mutations may be, on the whole, deleterious.
In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantagous in A/H3N2.
It seems possible that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
However, more work is necessary to investigate this hypothesis.

@@

@@

Antigenic, genetic and surveillance data is available as supplemental datasets \tbc{PREPARE}.

@@

We thank John McCauley for making NIMR vaccine strain selection reports publicly available.
We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.

>>>>

Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations progressing in a random direction.
However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
This result indicates that forward movement along antigenic dimension 1 is an advantageous trait and is promoted by natural selection \cite{Bedford12}.
The rate of diffusion of persistent lineages is substantially greater in A/H3N2 than in A/H1N1, B/Vic or B/Yam.
For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drifts along dimension 1.

%%% accumulation %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/accumulation}
	\caption{\textbf{Rate of antigenic diffusion of influenza lineages categorized by time back from sampling.}
	(A) Rate of diffusion in terms of antigenic units per year only along antigenic dimension 1.
	(B) Rate of diffusion in terms of antigenic units per year along both antigenic dimensions 1 and 2.
	In the diffusion model, the antigenic location of each virus sample can be calculated back through time.
	Here, when the $x$-axis is 1, we show the rate of diffusion 1 year back from from the date of sampling averaged across all virus samples.
	When the $x$-axis is close to 0, the $y$-axis shows the rate of diffusion near tips of the phylogeny, and when the $x$-axis is large, the $y$-axis shows the rate of diffusion along the trunk of the phylogeny.
	} 
	\label{accumulation} 
\end{figure}

@@

We suggest that this finding could be due to the pleiotropic effects associated with antigenic change, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantageous, while other antigenic mutations may be, on the whole, deleterious.
In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantageous in A/H3N2.
It seems possible that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
However, more work is necessary to investigate this hypothesis.

@@

In this study we provide a foundation for evolutionary antigenic cartography in which to simultaneously assess antigenic phenotype and patterns of antigenic evolution.
We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each influenza clade.
However, A/H3N2 shows substantially stronger signals of adaptive antigenic evolution, evolving faster in antigenic phenotype and showing more efficient conversion of transient antigenic polymorphism into fixed differences.
Consequently, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.

@@

Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.

@@

We thank John McCauley for making NIMR vaccine strain selection reports publicly available.
We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
We acknowledge the laboratories that provided sequences to EpiFlu database: Centers for Disease Control and Prevention (USA), Hospital Clinic de Barcelona, Institut National d'Hygi\`{e}ne du Maroc, National Institute for Medical Research (UK), National Institute of Infectious Diseases (Japan), Norwegian Institute of Public Health, Swedish Institute for Infectious Disease Control, WHO Chinese National Influenza Center, WHO Collaborating Centre for Reference and Research on Influenza (Australia).
